STAT 672 Final Project:\\ Stochastic Gradient Descent

### 2018-04-16 paper/paper.tex
more work

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}

@@

\title{STAT 672 Final Project: Stochastic Gradient Descent}

@@

The above proof is for GD. A rigorous proof is not given for why SGD also
converges to the optimal value. Informally, we can note that the above proof
says given infinite $t$ and appropriate $\gamma$, the iterative algorithm will
always converge to the optimal value. SGD is doing the same thing as GD, 
just with a single observation per iteration 
rather than the entire dataset per iteration. Since SGD is using less information
per iteration, we would expect the convergence to require more iterations. But
per (9) it is guaranteed to \textit{eventually} arrive at the optimal solution.
Put differently, the difference between GD and SGD must then primarily revolve
around convergence speed, not the final value that is converged to.

@@

We can visually depict SGD's ``noisier'' convergence in comparison to GD.

@@

The basic SGD algorithm has been extended in many different ways. The popularity of
the algorithm disallows a comprehensive or detailed treatment of all
development. This sub-section covers some of the more important and interesting extensions.

@@

As hinted at in our proof of convergence to $\hat{w}$, much depends on
an appropriately chosen step size $\gamma$. If $\gamma$ is too small, (S)GD may
take an excessively long time to run (since we are only moving a small distance
``down'' the gradient every iteration); if $\gamma$ is too large, then an
iteration's step may ``overshoot'' the critical point and prevent convergence.
An entire literature has developed around the best way to select $\gamma$. The
central insight that is common to most $\gamma$-selection methods is to treat it
as a time-indexed rather than fixed hyperparameter, i.e., $\gamma^{(t)}$ rather
than $\gamma$. Ideally we would like $\gamma^{(t)}$ to be large when far away
from the critical point (so as to increase speed of convergence) and to be small
when close to the critical point (so as to avoid overshooting).

Line search is one method for computing step size. As described in
\cite{boyd2004convex}, there are two main variants: \textit{exact} and
\textit{backtracking}. In exact line search, $\gamma$ is chosen to minimize $f$
along the ray $\{x + \gamma \Delta x\}$, where $\gamma \in \mathbb{R}^+)$ and
$\Delta x$ is the descent direction determined by SGD:
\begin{equation}
	\gamma = \argmin_{s \geq 0} f(x + s\Delta x)
\end{equation}
This obviously is an optimization problem in and of itself, and
so it can be computationally impractical to add this burden in addition to the
``main'' optimization problem we are trying to solve using SGD. For this reason,
\textit{backtracking} is an iterative approximation method that is
computationally lighter. The backtracking algorithm takes two hyper-parameters
$\alpha \in (0, 0.5)$ and $\beta \in (0, 1)$. It starts with an initial guess of
$s^{(0)} = 1$, and then for every iteration $t$, updates $s := \beta t$. The
algorithm stops when $f(x + s \Delta x) \leq f(x) + \alpha s \Delta f(x)^T
\Delta x$, and the final value of $s$ is taken as $\gamma$.

@@

In our set-up of SGD, we stipulated that the function to be minimized, $f$, is
convex. Although this is a necessary condition for some proofs of SGD's
correctness, SGD is widely used in applications---most prominently, neural
networks and deep
learning---where this assumption is not
always true (as covered in more detail in CHAPTER X). In such settings, it is
possible that there are multiple local minima. The challenge for SGD is
how to avoid getting ``stuck'' in these local ravines and continue iterating
until the true global minimum is achieved.

Momentum---also called acceleration---is a common technique for avoiding this. As outlined in
\cite{rumelhart1986general} and 
\cite{qian1999momentum}, we add a new acceleration term to the familiar GD
equation. Let $z^{(t+1)} = \alpha z^{(t)} + \nabla f(w^{(t)})$. GD with
momentum then is:

@@

Now, the weight vector obtained for the current timestep depends both on the
current gradient \textit{and the gradient of the previous time-step}.
Intuitively, without acceleration, SGD tends to oscillate within a ravine
because the gradient is dominated by the steep curves on either side of the
ravine; acceleration helps point the descent along the shallower curve leading
out of the ravine. \cite{nesterov} provides another popular implementation of
acceleration.
\cite{polyak1992acceleration}

@@

\subsubsection{Predictive Variance Reduction}

@@

SGD is commonly used in large-$n$, computationally demanding applications. Thus,
even though SGD is a computational improvement over batch GD, there has been
interest in whether SGD can be made even faster by parallelizing it.
\cite{zinkevich2010parallelized} present novel algorithms for doing so. The
actual algorithms are strikingly simple; their proof is highly technical and
omitted here.

The parallelization technique essentially is averaging. In line with previous
notation, suppose we have fixed learning rate $\gamma$,

Advantages

@@

\cite{shalev2011pegasos}
\subsection{SGD and Statistical Learning}
\cite{dal2015calibrating}

@@

\begin{equation}
	\bm{w}_{t+1} = \bm{w}_t - \gamma 
	\nabla L(Y_i, f_{\bm{w}_t}(\bm{X}_i))
\end{equation}
\section{Conclusion}

>>>>

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{graphicx}
\usepackage[linesnumbered, lined, boxed]{algorithm2e}
\RestyleAlgo{boxruled}

@@

\title{STAT 672 Final Project:\\ Stochastic Gradient Descent}

@@

The above proof is for GD. A rigorous proof is not given for why SGD also
converges to the optimal value. Informally, we can note that the above proof
says given infinite $t$ and appropriate $\gamma$, the iterative algorithm will
always converge to the optimal value. SGD is doing the same thing as GD, 
just with a single observation per iteration 
rather than the entire dataset per iteration. Since SGD is using less information
per iteration, we would expect the convergence to require more iterations. But
per (9) we expect it to \textit{eventually} arrive at the optimal solution.
If so, the difference between GD and SGD must then primarily revolve
around convergence speed, not the final value that is converged to.

@@

Figure 1 visually compares SGD's convergence to that of GD (and
mini-batch gradient descent, a variation addressed later in this
paper).\footnote{Image taken from www.towardsdatascience.com} The
path of SGD's convergence is very squiggly, reflecting the extra variation
introduced by only using a single observation per iteration. The path of GD's
convergence is much smoother, reflecting the use of all observations per
iteration.

\begin{figure}[h!]
	\centering
	\caption{Convergence}
\includegraphics[scale=0.3]{img}
\end{figure}

@@

The basic SGD algorithm has been extended in many different ways. The popularity of
the algorithm disallows a comprehensive or detailed treatment of all
developments. This sub-section covers some of the more important and interesting
ones.

@@

As hinted at in our proof of convergence to $\hat{w}$, much depends on
an appropriately chosen step size $\gamma$. If $\gamma$ is too small, gradient
descent may
take an excessively long time to run (since we are only moving a small distance
``down'' the gradient every iteration); if $\gamma$ is too large, then an
iteration's step may ``overshoot'' the critical point.
An entire literature has developed around the best way to select $\gamma$. The
central insight that is common to most $\gamma$-selection methods is to treat it
as a time-indexed rather than fixed hyperparameter, i.e., $\gamma^{(t)}$ rather
than $\gamma$. Ideally we would like $\gamma^{(t)}$ to be large when far away
from the critical point (so as to increase speed of convergence) and to be small
when close to the critical point (so as to avoid overshooting).

Line search is one method for computing step size. As described in
\cite{boyd2004convex}, there are two main variants: \textit{exact} and
\textit{backtracking}. In exact line search, $\gamma^{(t)}$ is chosen to minimize $f$
along the ray $\{x + \gamma^{(t)} \Delta x\}$, where $\gamma^{(t)} \in \mathbb{R}^+)$ and
$\Delta x$ is the descent direction determined by SGD:
\begin{equation}
	\gamma^{(t)} = \argmin_{s \geq 0} f(x + s\Delta x)
\end{equation}
This obviously is an optimization problem in and of itself, and
so it can be computationally impractical to add this burden in addition to the
``main'' optimization problem we are trying to solve using SGD. For this reason,
\textit{backtracking} is an iterative approximation method that is
computationally lighter. The backtracking algorithm takes two hyper-parameters
$\alpha \in (0, 0.5)$ and $\beta \in (0, 1)$. It starts with an initial guess of
$s^{(0)} = 1$, and then for every iteration $t$, updates $s := \beta t$. The
algorithm stops when $f(x + s \Delta x) \leq f(x) + \alpha s \Delta f(x)^T
\Delta x$, and the final value of $s$ is taken as $\gamma^{(t)}$.

@@

In our set-up of SGD, we stipulated that the function to be minimized, $f$, is
convex. Although this is a necessary condition for some proofs of SGD's
correctness, SGD is widely used in applications---most prominently, neural
networks and deep
learning---where this assumption is not
always true. In such settings, it is
possible that there are multiple local minima, and/or that there are areas where
the surface curves much more steeply in one dimension than another. The challenge for SGD is
how to avoid getting ``stuck'' in these local ravines and continue iterating
until the true global minimum is achieved. Even if there is no pathological
curvature, incorporating information about the local topology may result in more
efficient algorithms and faster convergence.

Momentum---also called acceleration---is a common modification of GD. As outlined in
\cite{rumelhart1986general} and 
\cite{qian1999momentum}, we add a new acceleration term to the familiar GD
equation. Let $z^{(t+1)} = \alpha z^{(t)} + \nabla f(w^{(t)})$. GD with
momentum then is:

@@

Now, the weight vector obtained for the current timestep depends both on the
current gradient \textit{and the update of the previous time-step}, with the
acceleration parameter $\alpha$ governing how much weight is given to the
previous time-step. 
This formulation leads to momentum: dimensions whose
gradients point in the same direction have proportionally greater updates, while
dimensions whose gradients exhibit strong curvature have proportionally
lesser updates. The effect is to help stop (S)GD from oscillating in ravines and
speed up convergence. There are many variations on acceleration, including
Nesterov accelerated gradient (NAG) (\cite{nesterov}), Adam
(\cite{kingma2014adam}), AdaGrad
(\cite{duchi2011adaptive}), and AdaDelta (\cite{zeiler2012adadelta}).

@@

Averaged SGD (ASGD)---to the author's knowledge first proposed in 
\cite{polyak1992acceleration}---is another variation. Typically, we consider
$\hat{w}$ (the optimal solution) to be that associated with the final iteration
of SGD, when some stop condition is met. For example, in a situation where we
run $N$ iterations before stopping, $\hat{w} = w^{(N)}$. In ASGD, we consider $\hat{w}$ 
the average of $w$ across all iterations, i.e.
\begin{equation}
	\hat{w} = \bar{w} = \frac{1}{N} \sum_{t=1}^N w^{(t)}
\end{equation}

The value proposition of ASGD is that sometimes SGD will oscillate around the
optimal solution. By taking the average across all updates, ASGD reduces this
noise and is likely to give a solution closer to the optimum. More sophisticated
versions of ASGD have been proposed in 
\cite{zhang2004solving}
and 
\cite{xu2011towards}.

\subsubsection{Mini-batch}

Mini-batch gradient descent (MGD) can be regarded as a mid-point between batch
and stochastic gradient descent in that it uses more than one but less than all of
the available observations. Suppose we have $n$ observations and divide them
into $m$ groups called mini-batches, each of equal size $b=\frac{n}{m}$. Now, at
every iteration $t$, randomly pick one of the mini-batches (all are equally
likely to be chosen). For that mini-batch:

\begin{equation}
	w^{(t+1)} = w^{(t)} - \gamma \frac{1}{b} \sum_{i=1}^b \nabla f(w^{(t)})
\end{equation}

MGD iterates with $t$ and converges to $\hat{w}$ in the normal manner. Every
update, a new $b$ is randomly chosen and the above algorithm iterated. 
\cite{dekel2012optimal} and \cite{li2014efficient} analyze MGD's speed of
convergence and suggest that has some advantageous qualities for
parallelization, a topic that is explained in more depth in the following
subsection.

@@

SGD is commonly used in large-$n$ applications. Thus,
even though SGD is a computational improvement over batch GD, there has been
interest in whether SGD can be made even faster by parallelizing it. As
background, typical computer applications are serial: problems are broken
down into a discrete series of instructions that are executed one after another
on a single processor. Parallel computing calls for breaking down problems into discrete parts that can
be solved concurrently and are executed simultaneously on multiple processors
(\cite{barney2010introduction}).
There are many paradigms of parallel computing but a useful high-level
distinction is between shared memory, in which all processors share a common
address space, and distributed memory, in which each processor or group of processors
has its own address space. The latter has become increasingly popular in
industry via Apache Hadoop and associated projects (e.g. Spark). Distributed
memory applications often require very complex computing strategies for how to combine the
work of many processors without a shared memory space into a single coherent answer.

This background helps to emphasize the novelty of the parallel
implementation of SGD proposed in \cite{zinkevich2010parallelized}. Though backed
by highly technical proofs, the actual algorithm is strikingly simple. Suppose we have $N$ processors. We seek to
conduct SGD to estimate $w$ across $n$ observations with fixed learning rate
$\gamma$ for some fixed number of steps $T$.

\medskip

\begin{algorithm}[H]
	Define $T = \frac{n}{N}$\;

	Randomly partition examples, giving $T$ to each machine;\

	\ForAll{$i \in \{1 \ldots N\}$}{
		Randomly shuffle data on machine $i$\;

		Initialize $w^{i, (0)}=0$\;

		\ForAll{$t \in \{1 \ldots T\}$}{

	$w^{i, (t+1)} = w^{i, (t)} - \gamma \nabla f(w^{i, (t)})$ \;
			}
		}
	Aggregate from all computers: $\bar{w} = \frac{1}{N} \sum_{i=1}^{N}
	w^{i, (T)}$ \;

	\Return{$\bar{w}$}

	\caption{Parallel SGD}
\end{algorithm}

\medskip

Zinkevich et al. show that this algorithm has a number of desirable properties.
It requires no communication between machines until the end; asymptotically, the
error approaches zero; and, the amount of time required is independent of the number of
examples (though this is an inherent property of SGD and is not unique to their
algorithm). Further improvements have been proposed: e.g., the Hogwild algorithm
presented in \cite{recht2011hogwild} deals with locking issues in parallel
computation of SGD.

@@

In theory, SGD is agnostic to $f$ and $w$ as long as they meet required
assumptions, and so can be used in many different areas of optimization. In
practice, SGD is overwhelmingly popular in the statistical learning
field for estimating weight coefficients in predictive models. 
This section lays out some reasons for SGD's popularity; conducts a case study
using SGD; and identifies notable real-world applications of SGD.

\subsection{SGD and Statistical Learning}
\cite{shalev2011pegasos}

@@

\begin{equation}
	\bm{w}_{t+1} = \bm{w}_t - \gamma 
	\nabla L(Y_i, f_{\bm{w}_t}(\bm{X}_i))
\end{equation}

\subsection{Case Study}

\cite{dal2015calibrating}

\subsection{Real-World Applications}

If a Silicon Valley press release uses words like ``artificial intelligence'',
``machine learning'', and ``big data'' to describe a new product or service, it very likely
uses SGD in some way. Such is the extent of SGD's popularity. 
Below is a list of particularly prominent examples of SGD's practical
applications.

\begin{itemize}
	\item \textbf{ImageNet} is a dataset consisting of millions
		of high-resolution, labeled images. The
		ImageNet Large Scale Visual Recognition Competition (ILSVRC) is
		an annual contest in which competing teams train models for
		automatic image classification. The submission of \cite{krizhevsky2012imagenet} 
		for the 2012 ILSVRC used convolutional neural
		nets to achieve dramatic reduction in error rates and is
		one of the most cited machine learning papers of all time. Per the aforementioned paper: ``We
		trained our models using stochastic gradient descent with a
		batch size of 128 examples, momentum of 0.9, and weight decay of
		0.00005.''
	\item Google's \textbf{AlphaGo} software. AlphoGo applies deep learning to the
		board game of Go and has received extensive attention for its
		ability to defeat even top human players (e.g., twice making the
		cover of \textit{Nature} with \cite{silver2016mastering} and
		\cite{silver2017mastering}). Google's slides from the 2016
		International Conference on Machine Learning (ICML) indicate
		that SGD is used in every stage of the
		model: supervised learning of policy networks, reinforcement
		learning of policy networks, and reinforcement learning of value
		networks.
	\item The \textbf{Netflix Grand Prize} was a competition hosted by
		Netflix in which teams competed to develop models for
		using Netflix data to make better movie recommendations
		(\cite{bennett2007netflix}). Most of the top entries used
		SGD for estimating model coefficients; for example,
		\cite{koren2009bellkor} states that ``in order to learn the
		involved parameters$\ldots$ learning is done by a stochastic
		gradient descent algorithm running for 30 iterations.'' 
	\item Google's \textbf{word2vec} natural language processing (NLP)
		software powers Google Translate, among other applications.
		\cite{mikolov2013distributed} and \cite{rong2014word2vec}
		discuss the continuous bag of words (CBOW) model underlying
		word2vec and how SGD is used to estimate coefficients
		in the model.
\end{itemize}

\section{Conclusion}
